Ecological management of stochastic systems with long transients

### 2019-12-30 manuscript/manuscript.Rmd
WIP writing writing

library(ggplot2)
library(ggthemes)
library(ggtext) # devtools::install_github("clauswilke/ggtext")
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
pal <- solarized_pal()(3)
```

@@

Where $X_t$ is the population density of the focal species (the beetle, in our case).  The growth  rate of the species is determined by an intrinsic growth rate $r$ and carrying capacity $K$. The species is under mortality pressure from another species (the natural defences of trees, in our context) which has an S curve saturation which reaches its half maximum at $H$ and maximum rate of $a$.  $Q$ determines the steepness of the saturating response. The dynamics are subject to environmental white noise proportional to the population size, $\xi_t X_t$ [@Boettiger2018]. Model dynamics are illustrated in Figure 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.

@@

```{r fig_model, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$. Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.  By changing parameter $a$, the system goes from bistable (red, $a = 28$) to a ghost attractor (blue, $a = 27.2$). For both curves, $r = 0.8$, $K = 153$, $Q = 2$, $H = 20$."}
c(`ghost attractor` = 27.2,           
  `bistable attractor` = 28) %>%
  map_dfr(function(a) tibble(x = states, f = may(a)(x,0) - x, a = a), .id = "scenario") %>% 
  ggplot(aes(x, f)) + 
    geom_line(aes(lty = scenario, col = scenario), lwd = 1) +
    geom_hline(aes(yintercept = 0)) + 
    labs(y = bquote(X[t+1] - X[t]), x = bquote(X[t])) + 
    scale_color_solarized()

@@

Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.  Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Figure 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of what @Hastings2018 has dubbed a "Ghost Attractor" -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed.  This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

This same model structure can thus produce rather different dynamics due to only a small change in a single parameter, as illustrated in the ensemble simulations shown in Figure 2. Each ensemble shows 50 replicate trajectories starting with the same initial condition, under parameters corresponding either to the bistable system (panel A) or the ghost attractor (panel B). Because the model is stochastic, trajectories under the bistable system have some probability of escaping the lower attractor after a sufficient interval.  Thus while it is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastically escaping stable system.

@@

```{r fig_sims}
fig1_sims %>%
  ggplot(aes(time, state, group = interaction(rep, scenario), col= scenario)) + 
  geom_line(alpha=0.3) + scale_color_solarized() + facet_wrap(~scenario)
```

@@

In the remaining sections, I explore two different approaches to addressing the problem of management under a potential ghost attractor.  The first considers the decision-theoretic optimal policy given uncertainty

@@

@@

First, I will consider the case that in which the manager has been able to accurately infer the dynamics (i.e. correctly estimate the parameters of Eq~\eqref{1}), and assuming the model structure postulated \eqref{1} is correct.  The optimal strategy determined by SDP can be represented as a "policy" function $p(X)$, indicating for each possible state $X$ what the optimal action $\hat A = p(X)$ would be.  Figure 3 displays the numerically computed optimal policy for the system with only a transient.  Though the models differ by only a small amount in a single parameter, because one model corresponds to a system with only a ghost attractor while the other has a stable state attractor, the resulting optimal policy solutions differ significantly.  Intuitively, in the system with stable state dynamics, the manager does not implement any control at low levels, where the presence of the lower stable state will constrain the beetle population to acceptable levels.  Once that lower stable state becomes only a ghost attractor, the optimal solution targets elimination of the beetle even at low densities.  Consequently, the optimal response becomes tightly dependent on our knowledge of the dynamics.

@@

```{r fig_policy}

policy_plot %>%
  ggplot(aes(state, action, color = model)) + 
  geom_point(alpha=0.6) + 
  scale_color_solarized()
```

@@

The transition matrix is an expression of conditional probability $P(X_{t+1} | X_t,  A_t)$, the probability that a system in state $X_t$ will transition to state $X_{t+1}$ in the next time interval if the manager takes action $A_t$.  If we have $n$ possible states and $m$ possible actions, we have $m \phantom{\cdot} n \times n$ matrices for the possible transitions.  The transistion matrix also depends on our knowledge or estimates of the parameters of \eqref{model}. We can represent uncertainty over the models and their parameters as uncertainties as conditional on the model, $M_i$, so that the transistion matrix must then be marganlized over the set of possible models: $$P(X_{t+1} | X_t,  A_t) = \sum_{i} P(X_{t+1} | X_t,  A_t, M_i) P(M_i)$$

@@

```{r eval=FALSE}
## Figure not shown -- intermediate prior gives an intermediate policy
tibble(state = states,
       action = actions[fifty_fifty$policy])  %>% 
  ggplot(aes(state, action)) + geom_point() + 
  geom_vline(aes(xintercept = states[x0]))
```

@@

## Rather stochastic in realization space, should consider ensembles instead...
reps <- 10
Tmax <- 60
set.seed(123456)
sim_ghost <- map_dfr(1:reps, function(i)
                   mdp_planning(P_ghost, reward, discount, 
                   policy = soln_ghost$policy,
                   x0 = x0, Tmax = Tmax), .id = "reps")
sim_bistable <- map_dfr(1:reps, function(i)
                   mdp_planning(P_ghost, reward, discount, 
                   policy = soln_bistable$policy,
                   x0 = x0, Tmax = Tmax), .id = "reps")
sim_fiftyfifty <-  map_dfr(1:reps, function(i)
                   mdp_planning(P_ghost, reward, discount, 
                   policy = fifty_fifty$policy,
                   x0 = x0, Tmax = Tmax), .id = "reps")
sim_seventythirty <- map_dfr(1:reps, function(i)
                   mdp_planning(P_ghost, reward, discount, 
                   policy = seventy_thirty$policy,
                   x0 = x0, Tmax = Tmax), .id = "reps")

sims_by_prior <- 
 bind_rows(ghost = sim_ghost, 
           bistable = sim_bistable, 
           fifty_fifty = sim_fiftyfifty, 
           seventy_thirty = sim_seventythirty, 
           .id = "prior")

@@

```{r fig_sims_by_prior}
sims_by_prior <- read_csv("manuscript_cache/sims_by_prior.csv")

sims_by_prior %>% 
  mutate(state = states[state], action = actions[action]) %>% 
  select(-obs,-value) %>%
  pivot_longer(c(-time, -prior, -reps)) %>%
  ggplot(aes(time, value, group = interaction(name, reps), col=name)) + 
  geom_path(alpha=0.5) + facet_wrap(~prior)
```

@@

This probability thus depends on our prior belief that model $i$ is true, $P(M_i)$.  Not only does this allow our manager to make her decision in to reflect uncertainty over models, but it also suggests that having made that decision and observed the outcomes, the manager can update her belief over possible models according to Bayes Rule,

$$P(M_i | X_{t+1} ) = \frac{P(X_{t+1} | M_i ) P(M_i)}{P(X_{t+1})}$$
that is, the optimal manager not only plans over the uncertainty, but also learns which model is most likely to be correct over the course of management.

```{r mdp_learning}
set.seed(12345)
Tmax <- 50
learning <- mdp_learning(list(P_ghost, P_bistable), reward, discount, 
                   model_prior = c(0.5, 0.5),
                   x0 = x0, Tmax = Tmax,
                  true_transition = P_ghost,
                  max_iter = 1e4, epsilon = 1e-2)

@@

```{r fig_learning_sim}
 learning$df %>% 
  select(-value) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, states[state], color = series)) + geom_line()
```

@@

# Methods

>>>>

library(patchwork)
library(ggplot2)
library(ggthemes)
library(ggtext) # devtools::install_github("clauswilke/ggtext")
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
pal <- solarized_pal()(7)
scale_colour_discrete <- function(...) scale_colour_solarized()
scale_fill_discrete <- function(...) scale_fill_solarized()
```

@@

Where $X_t$ is the population density of the focal species (the beetle, in our case).  The growth  rate of the species is determined by an intrinsic growth rate $r$ and carrying capacity $K$. The species is under mortality pressure from another species (the natural defences of trees, in our context) which has an S curve saturation which reaches its half maximum at $H$ and maximum rate of $a$.  $Q$ determines the steepness of the saturating response. The dynamics are subject to environmental white noise proportional to the population size, $\xi_t X_t$ [@Boettiger2018]. Model dynamics are illustrated in Fig 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.

@@

```{r fig_model, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$. Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.  By changing parameter $a$, the system goes from bistable (red, $a = 28$) to a ghost attractor (blue, $a = 27.2$). For both curves, $r = 0.8$, $K = 153$, $Q = 2$, $H = 20$."}

## Fig 1

c(`ghost attractor` = 27.2,           
  `bistable attractor` = 28) %>%
  map_dfr(function(a) tibble(x = states, f = may(a)(x,0) - x, a = a), .id = "scenario") %>% 
  ggplot(aes(x, f)) + 
    geom_line(aes(lty = scenario, col = scenario), lwd = 1) +
    geom_hline(aes(yintercept = 0)) + 
    labs(y = bquote(X[t+1] - X[t]), x = bquote(X[t])) + 
    scale_color_solarized()

@@

Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.  Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of what @Hastings2018 has dubbed a "Ghost Attractor" -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed.  This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

This same model structure can thus produce rather different dynamics due to only a small change in a single parameter, as illustrated in the ensemble simulations shown in Fig 2. Each ensemble shows 50 replicate trajectories starting with the same initial condition, under parameters corresponding either to the bistable system (panel A) or the ghost attractor (panel B). Because the model is stochastic, trajectories under the bistable system have some probability of escaping the lower attractor after a sufficient interval.  Thus while it is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastically escaping stable system.

@@

```{r fig_sims}
# Fig 2
fig1_sims %>%
  ggplot(aes(time, state, group = interaction(rep, scenario), col= scenario)) + 
  geom_line(alpha=0.3) + scale_color_solarized() + facet_wrap(~scenario)
```

@@

In the remaining sections, we will explore two different approaches to addressing the problem of management under a potential ghost attractor. First, we will examine the decision-theoretic optimal policy for each stochastic model separately, illustrating that the transient system does indeed lead to a different management strategy than the alternate stable state system.  We will then examine the optimal solution when the manager is uncertain which of the two models is the true.  Finally, we will examine an active adaptive management approach, in which the manager updates the probability that each model is true following each iteration of decision and observation.

@@

The transition matrix is an expression of conditional probability $P(X_{t+1} | X_t,  A_t)$, the probability that a system in state $X_t$ will transition to state $X_{t+1}$ in the next time interval if the manager takes action $A_t$.  If we have $n$ possible states and $m$ possible actions, we have $m \phantom{\cdot} n \times n$ matrices for the possible transitions. The transistion matrix also depends on our knowledge or estimates of the parameters of \eqref{model}. For simplicity, the examples here have assumed a grid of state space values as he integers between 0 to 120, and an equivalently fine grid of 121 possible actions (control levels).  This provides sufficient approximation to a continuous state space and action space, so that results are not sensitive to the details of this discritization.

@@

First, let us consider the case that in which the manager has been able to accurately infer the dynamics (i.e. correctly estimate the parameters of Eq~\eqref{1}), and assuming the model structure postulated \eqref{1} is correct.  The optimal strategy determined by SDP can be represented as a "policy" function $p(X)$, indicating for each possible state $X$ what the optimal action $\hat A = p(X)$ would be.  Fig 3 displays the numerically computed optimal policy for the system with only a transient.  Though the models differ by only a small amount in a single parameter, because one model corresponds to a system with only a ghost attractor while the other has a stable state attractor, the resulting optimal policy solutions differ significantly.  Intuitively, in the system with stable state dynamics, the manager does not implement any control at low levels, where the presence of the lower stable state will constrain the beetle population to acceptable levels.  Once that lower stable state becomes only a ghost attractor, the optimal solution targets elimination of the beetle even at low densities.  Consequently, the optimal response becomes tightly dependent on our knowledge of the dynamics.

@@

```{r fig_policy, fig.cap="Optimal policy functions under each model. Parameter values for utility function used here are $c = 1$, $d = 0.5$, $X_0 = 50$, $\delta = 0.999$, parameters for the transition models are as given in Fig 1. "}
## Fig 3

policy_plot %>%
  ggplot(aes(state, action, color = model)) + 
  geom_point(alpha=0.6)
```

```{r}
reps <- 20
Tmax <- 200
set.seed(12345)
sim_ghost <- map_dfr(1:reps, function(i)
                   mdp_planning(P_ghost, reward, discount, 
                   policy = soln_ghost$policy,
                   x0 = x0, Tmax = Tmax), .id = "reps")
sim_bistable <- map_dfr(1:reps, function(i)
                   mdp_planning(P_ghost, reward, discount, 
                   policy = soln_bistable$policy,
                   x0 = x0, Tmax = Tmax), .id = "reps")

fixed_sims <- 
 bind_rows(ghost = sim_ghost, 
           bistable = sim_bistable, 
           .id = "prior") %>% 
  mutate(state = states[state], 
         action = actions[action])

fixed_sims %>% write_csv("manuscript_cache/fixed_sims.csv")
```

We can get a better idea of the consequences of mistaking the long transient dynamics for stable state dynamics by examining the dynamics of a system when it is actively managed according to the policy function for the wrong model.  It is important to note that even when the policy wrongly assumes the system has an endemic stable state (e.g. small beetle population density), this does not necessarily mean that the manager will never intervene and the dynamics will reach the high stable state unchecked. As we see in the policy function, at low densities, the alternative stable state policy does indeed choose to do nothing, since it expects the density to remain around an acceptable endemic level.  However, once the densities get high enough (the system escapes the ghost attractor), the manager intervenes -- but this intervention is only enough to restore the system to the endemic state.  This results in a pattern of perpetual interventions. Simulations of 10 independent realizations of a ghost-attractor system being managed with the optimal policy of the alternative stable state model (Fig 4, upper panel).  In contrast, had the manager known the system no longer supported the endemic stable state but had only a ghost attractor, the optimal policy is a costly up-front intervention which eliminates the beetle entirely (Fig 4, lower panel).  A comparison between the large up front cost of the optimally managed realizations to the ever rising costs of the mistaken case can be seen in Fig 5, shown in terms of net present value (i.e. cumulative costs, where future costs have been discounted by rate $\delta$.)  The examples shown here use a relatively small discount rate of 0.999 per unit time -- under sufficiently agressive discounting this large up-front investment would not be optimal.  (Note that we use the term "small" meaning relative to the timescale shown, e.g. the value at time 200 is worth $0.999^{200} \approx 81.86\%$ of its starting value. As the units of time here correspond to the generation time of the beetle, this shouldn't be confused with an *annual* discount rate.)

```{r fig_sims_by_prior}
# Fig 4
fixed_sims <- read_csv("manuscript_cache/fixed_sims.csv")

fixed_sims %>% select(state, action, time, prior, reps) %>%
  pivot_longer(c(-time, -prior, -reps), names_to = "line", values_to  = "state") %>%
  filter(time < 100) %>%
  ggplot(aes(time, state, group = interaction(line, reps), col=line)) + 
  geom_line(alpha=0.5) + 
  facet_wrap(~prior, ncol=1)
```

```{r}
costs <- fixed_sims %>% group_by(time, prior)  %>% 
  summarise(cost = mean(value)) %>%
  group_by(prior) %>%
  mutate(cost = abs(cumsum(cost * .999 ^ time))) %>%
  rename(model = prior)

costs %>%
  ggplot(aes(time, cost, ymin = 0, ymax = cost, fill=model)) +
  geom_ribbon(alpha = 0.5) +
  ylab("Cost (net present value)") + xlab("time (days)")
```

```{r}
# compare values
```

@@

Rather than having to completely assume a given model, it is possible to reflect our uncertainty about which model is correct in our calculation of the optimal policy.   We can represent uncertainty over both the models and their given parameters conditionally: so that the transistion matrix is marginalized over the set $M_i$ of possible models:

$$P(X_{t+1} | X_t,  A_t) = \sum_{i} P(X_{t+1} | X_t,  A_t, M_i) P(M_i)$$

This requires we can assign probabilities that each model is true, which I will refer to as 'prior probabilities' for reasons that will be clear in the next section. In this way, our uncertainty about the future state reflects both the inherent uncertainty of the stochastic process as well as our ignorance of the particular model. For simplicity I will consider only the two spcific models we have worked with so far, though this same approach could easily be extended to a suite of possible models.  If the two models are considered equally likely, the optimal policy in the case of our examples looks like an only slightly more agressive version of the policy for the bistable system, as seen in the replicate simulations shown in Fig 6.  This policy eliminates the beetles in a few of the replicate simulations where the population dips enough to make that decision economical, but otherwise settles for persistent management, as shown in the upper panel.  However, we can also see that absolute certainty is not required for definitive action -- once we are sufficiently sure the system is a ghost attractor, e.g. 70% probability as shown in the lower panel of Fig 6, the optimal solution eliminates the pest immediately on all replicates.  Not surprisingly, the 50-50 case achieves a long run net present value that is intermediate to the optimal solution and the mistaken solution considered in Fig 5.

@@

```{r eval=FALSE}
## Fig not shown -- intermediate prior gives an intermediate policy
tibble(state = states,
       action = actions[fifty_fifty$policy])  %>% 
  ggplot(aes(state, action)) + geom_point() + 
  geom_vline(aes(xintercept = states[x0]))
```

@@

sim_fiftyfifty <-  map_dfr(1:reps, function(i)
                   mdp_planning(P_ghost, reward, discount, 
                   policy = fifty_fifty$policy,
                   x0 = x0, Tmax = Tmax), .id = "reps")
sim_seventythirty <- map_dfr(1:reps, function(i)
                   mdp_planning(P_ghost, reward, discount, 
                   policy = seventy_thirty$policy,
                   x0 = x0, Tmax = Tmax), .id = "reps")

sims_by_prior <- 
 bind_rows(fifty_fifty = sim_fiftyfifty, 
           seventy_thirty = sim_seventythirty, 
           .id = "prior")  %>% 
  mutate(state = states[state], 
         action = actions[action])

@@

```{r fig_sims_by_prior}
## Fig 6

sims_by_prior <- read_csv("manuscript_cache/sims_by_prior.csv")

sims_by_prior  %>% select(state, action, time, prior, reps) %>%
  pivot_longer(c(-time, -prior, -reps), names_to = "line", values_to  = "state") %>%
  filter(time < 100) %>%
  ggplot(aes(time, state, group = interaction(line, reps), col=line)) + 
  geom_line(alpha=0.5) + 
  facet_wrap(~prior, ncol=1)
```

@@

In the scenarios we have just considered, the manager continues to apply the same optimal decision rule over the entire management period.  This does not, of course, mean that the management response is completely static -- in fact, the manager selects a different the intervention level at each timestep in response to updated information about the current beetle population size -- but the manager does not use that information to revisit the underlying assumptions about the appropriate model. However, each additional observation potentially provides further information about which model is correct.  After enough attempts to restore the population to a stable endemic state only to have another outbreak follow, we might expect our manager to realize that the endemic state just isn't as stable as originally thought.  More formally, we might assume the manager would update that initial prior belief over the possible models according to Bayes Rule,

$$P(M_i | X_{t+1} ) = \frac{P(X_{t+1} | M_i ) P(M_i)}{P(X_{t+1})}$$
In other words, our optimal manager not only plans over the uncertainty, but also learns which model is most likely to be correct over the course of management.  This approach corresponds to Carl Walters' original notion of "adaptive management" [@Walters1978]. (Note that the term is now used much more loosely, and would commonly include the kind of decision rule shown previously where the decision is adjusted based on future observations.).  This approach is considerably more computationally intensive, since the stochastic dynamic programming problem must be solved again at each time step after the model weights have been updated. Fig 7 illustrates a typical outcome of adaptive management starting with the uniform prior (equal weights on both models): a short period of modest intervention while the manager learns which model is more likely to be correct, followed by a more decisive intervention once the long transient becomes sufficiently likely.  Of course, this defensive action also eliminates the potential for future learning, so the manager can never be completely sure this action was necessary, as shown by the model probabilities in Fig 8.  This represents a form of precautionary principle already suggested by the purely planning approach considered in the previous section -- the manager need not be certain about the correct model, but merely reach a certain threshold of probability.

```{r mdp_learning}
# Fig 7
set.seed(12345)
Tmax <- 50
learning <- mdp_learning(list(P_ghost, P_bistable), reward, discount, 
                   model_prior = c(0.5, 0.5),
                   x0 = x0, Tmax = Tmax,
                  true_transition = P_ghost,
                  max_iter = 1e4, epsilon = 1e-2)

@@

```{r fig_learning_sim}
# Fig 8
 learning$df %>% 
  select(-value) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, states[state], color = series)) + geom_line()
```

@@

# Discussion

Long transients. Equilibrium.

Note that this approach can equally well consider an arbitrary suite of models (i.e. varying the parameter $a$ over a range of values, as well as to models with completely different structure).

Not considered:

- partially observed optimal control
- non-stationary dynamics e.g. in which

Implications:

- Long transients can be misleading -- e.g. mistaking a ghost attractor for a stable state.
- These mistakes can be significant, since long transients require qualitatively different management strategies.
- Nevertheless, thes

# Numerical Methods
